### 2019-03-19 proposal/proposal.tex
proposal latex

\documentclass{article}
\usepackage{hyperref}

\usepackage[letterpaper, margin=1in]{geometry}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0pt}

\title{6.890 Spring 2019 Project Proposal \\ LHYRA: Learned HYbrid Recursive Algorithms}

\author{Josh Gruenstein \and Lior Hirschfeld \and Ben Spector}

\date{\today}

\begin{document}

\maketitle

\subsection*{Introduction}

We propose a framework for automatically learning recursive algorithms and data structures for arbitrary data distributions and optimization criteria. Our ideas are loosely inspired by Prof. Kraska’s work on learned B-Trees and the Recursive Model Index (RMI); we conceive of this project as a sort of generalization of this idea to most recursive data-structures and algorithms.

We observe that there are diverse problems in computer science that can be solved by recursive structures\footnote{Not all algorithms used in our framework need be recursive; in some cases it may be best to solve a large base-case directly (for example, using insertion-sort for small nearly-sorted lists), and we recommend providing all the options to LHYRA and letting it sort out everything.}, from sorting to hierarchical pathfinding for online applications. Often though, these solutions are operating on a distribution which is very limited compared to the overall space of solutions, or the solutions are heuristics because finding perfect solutions is either intractable or impossible due to imperfect information. With such problems, there is usually some set of possible algorithms which produce solutions, but it is rare that a single solver is optimal for all possible distributions and objective functions: for example, insertion sort is preferable to merge sort in memory, or in time for nearly-sorted lists, but inferior in time for the random case.

Practically, what usually happens is that an expert programmer writes a number of these algorithms, benchmarks them against each other, and chooses the one which appears to perform the best. A clear example of this is in the C++ STL, in which the standard algorithm is Introsort, which calls Quicksort until hitting a certain recursion depth, where it switches to Heapsort. We observe that it would be better that these recursive structures would adapt to their distributions and objective functions, so that not only could they lessen the workload for the expert, but also better explore the potential synergies of the possible algorithms.

We therefore propose the Learned HYbrid Recursive Algorithms framework, or LHYRA. We give a more precise proposal below, but the essential idea is to automate the search of these learned hybrid models given data representing a distribution, candidate solvers, and a cost function. LHYRA will automatically search through the possible candidates (and potentially their hyperparameters, too) in order to determine the optimal algorithm for the task at hand.

\subsection*{Framework}

A LHYRA instance is described by the following components:

\begin{itemize}
	\item \textbf{Bag of solvers $S$.} Each solver $s_i \in S$ is equipped to solve an instance of the problem at hand, potentially by calling the LHYRA instance on subproblems. Each solver $s_i$ may also have a set of hyperparameters $h_ij$.

	\item \textbf{Data Generator / Training Data Set $D$.} In the training phase the problems in this set / generated by this generator are used by the optimizer to learn how to select a solver $s_i$ from S given problem features.

	\item \textbf{Feature extractor $F(x)$.} Our goal is to learn how to select the ideal solver based on a given input, thus we need a nice way of extracting useful features about an input. For example, a feature extractor for a sorting LHYRA instance may return a vector containing the length of the list, number of unique elements, and how sorted it already is. Defaults to a vector of length 1 containing the depth of the recursion.

	\item \textbf{Optimizer $O$.} Given a feature extractor, a training data set, and a bag of solvers, the optimizer learns how to select and parametrize a solver given a vector of problem features.  Optimizers could be implemented at varying degrees of complexity, but we imagine simple solutions such as SVMs to be competitive, and more complex solutions such as neural networks to start to saturate the gains one could achieve from a better optimizer module (at the cost of computation time). Optimizers should default to minimizing computation time and/or memory, but can take an arbitrary cost function $C(x,y)$ that produces some cost to be minimized given a problem and solution. 
\end{itemize}

Once initialized, a call to a LHYRA instance solves a problem x through the following steps:

\begin{enumerate}
	\item Compute $F(x)$.
	\item Pass $F(x)$ into the solver selector learned by the optimizer, which returns a parametrized solver $s_ih$.
	\item Compute $s_ih(x)$.
	\begin{enumerate}
		\item For each subproblem $y_i$ identified by $s_ih$, repeat the above process.
		\item Combine the subproblems and return a result.
	\end{enumerate}
\end{enumerate}

Ideally, the solver selector and feature extractor would be fast enough for the benefits gained by ideal solver selection to outweigh the additional overhead. However, after implementing LHYRA, we may discover a number of possible outcomes:

\begin{enumerate}
	\item LHYRA instances result in observable performance improvements in a wide array of real-world tasks.
	\item LHYRA instances result in observable performance improvements for a few specific tasks.
	\item LHYRA instances do not result in any observable performance improvements.
\end{enumerate}

Regardless of the results, our expansive benchmarking will allow us to determine which data structures and applications are most improved by our use of multiple solvers (ignoring the time spend in the decision making process). This information could lead to future advancements both in application specific tasks (where the behavior of our solver selector could be manually replicated) and in generalized cases, as more efficient solver selectors / feature extractors are found.

\subsection*{Optimizer Design}

One interesting component of learning hybrid recursive algorithms is developing generic classifiers (and their accompanying optimizers) for recursive algorithms.  In LHYRA, this whole problem is encapsulated by the Optimizer abstraction.

If one were to think of the most naive possible way to do this, you could imagine randomly sampling the bag of solvers (including their hyperparameters) and the training data to build a table of (feature set, solver) $\rightarrow$ cost. One could then perform a variant of nearest neighbor on that table given an input $x$ to find the solver with the lowest cost.

However, this approach is rife with complication. First, it is difficult using this framework to identify relationships between solvers (for example, if one solver is very good at solving its own subproblems). It would even be difficult to individually measure solver performance, as a fast solver that calls big subproblems might actually be less efficient than a slow solver that calls small subproblems. Even if one tries to get around this issue by averaging subproblem approaches, one still runs into the above issue of failing to model more complex relationships between solvers and the problems they best perform on.

Instead, we plan to model our optimization as a sort of reinforcement learning problem.  We start with a randomly initialized solver picker $P(x) \rightarrow s_i$, and use it to evaluate LHYRA on an example from the training set.  We record the time to perform this entire process, and treat that as a reward signal.  We can then either use fancy methods from reinforcement learning (such as Q-Learning) to update our classifier that have the potential to understand these more complex solver relationships, or even just perform simulated annealing on $P$'s parameters.  Regardless, this provides a far more powerful framework for designing optimizers.

However, we do note a specific subcase in which the above is not ideal: that of intentional overfitting.  For applications like RMI where the training data is the evaluation data, we should instead sample the entire solution space, and build a hashmap that provides $O(1)$ evaluation by looking up subproblems (this is possible, as we’ve already seen them before).  Thus we expect to build at least two Optimizers: at least one using the RL method above, and one using this $O(1)$ method.

\subsection*{Potential Applications}

Many problems and data structures in computer science are naturally recursive, and thus we expect no shortage of potential applications to benchmark LHYRA. Below are some of the most obvious examples we can think of where we might expect performance gains over a non-learned solution.

\begin{itemize}
	\item \textbf{B-Trees.} If possible, we would like to re-implement Kraska’s B-Trees in LHYRA to demonstrate how LHYRA is a true generalization. As LHYRA is partially modelled after RMI, we expect this to be a fairly straightforward adaptation.
	\item \textbf{Sorting.} We’d like to show how LHYRA can adapt to different distributions in choosing the optimal sorting method. We’d also like to show that LHYRA can also adapt to quickly sort approximately, and utilize processor parallelism where available.
	\item \textbf{Hierarchical Pathfinding.} Traditional pathfinding algorithms waste substantial amounts of computation and have difficulty adjusting to dynamic, online environments. The method of choice for hierarchical pathfinding is likely highly dependent on the exact optimization criteria, graph connectivity, and dynamism of the environment. This is a place where LHYRA may be invaluable, and also has clear real-world applications.
	\item \textbf{Matrix Multiplication.} LHYRA should be able to adaptively select naive matrix multiplication, naive divide and conquer, Strassen’s algorithm, or Coppersmith–Winograd. This would be particularly useful for machine learning.
	Nearest Neighbor: Given high-dimensional data, find the nearest neighbor. Examples of methods for this could be Locality sensitive hashing, Greedy search in proximity neighborhoods, and naive search.

\end{itemize}

\subsection*{Practical Notes}

This is a project with a lot of breadth, flexibility, and engineering effort, which we believe warrants a three person group. We intend to develop LHYRA in Python for the sake of implementation speed, as we can still demonstrate performance improvements over naive Python implementations.

We expect that in addition to building implementations for the above applications (each with its own bag of solvers, etc), we’ll experiment with different optimizers, classifiers, and additional applications. We’d also like to open source our codebase, and write sufficient documentation that future programmers may be able to use LHYRA for their own applications. We've setup a publically accessible repository here:

\begin{center}
\url{https://github.com/joshuagruenstein/lhyra}
\end{center}

\end{document}

### 2019-03-19 proposal/proposal.tex
tex typo fix

@@
 	\item \textbf{Bag of solvers $S$.} Each solver $s_i \in S$ is equipped to solve an instance of the problem at hand, potentially by calling the LHYRA instance on subproblems.
-Each solver $s_i$ may also have a set of hyperparameters $h_ij$.
+Each solver $s_i$ may also have a set of hyperparameters $h_{ij}$.
 
@@
 	\item Compute $F(x)$.
-	\item Pass $F(x)$ into the solver selector learned by the optimizer, which returns a parametrized solver $s_ih$.
-	\item Compute $s_ih(x)$.
+	\item Pass $F(x)$ into the solver selector learned by the optimizer, which returns a parametrized solver $s_{ih}$.
+	\item Compute $s_{ih}(x)$.
 	\begin{enumerate}
-		\item For each subproblem $y_i$ identified by $s_ih$, repeat the above process.
+		\item For each subproblem $y_i$ identified by $s_{ih}$, repeat the above process.
 		\item Combine the subproblems and return a result.

### 2019-05-13 c++/compile.sh
worksssssss

@@ -1,2 +1,2 @@
-g++ -I/home/sydriax/gsl/include -c lhyra_test.cpp -std=c++11
-g++ -L/home/sydriax/gsl/lib lhyra_test.o -lgsl -lgslcblas -lm -o lhyra_test -std=c++11
+g++ -I/home/sydriax/gsl/include -c sort_test.cpp -std=c++11
+g++ -L/home/sydriax/gsl/lib sort_test.o -lgsl -lgslcblas -lm -o sort_test -std=c++11

### 2019-05-13 c++/lhyra.hpp

@@ -40,44 +40,58 @@ public:
 }*/
 
 template<typename T, typename U>
-using Solver = std::function<U(T, std::function<U(T)>)>;
+using Solver = std::function<U(const T &, std::function<U(const T &)> &)>;
 
-template<typename T>
+template<typename T, unsigned int SIZE>
 class FeatureExtractor {
 public:
-    virtual std::vector<double> operator()(const T & t) = 0;
+    virtual std::array<double, SIZE> operator()(const T & t) = 0;
 };
 
 // Early declaration for the compiler
-template<typename T, typename U> class Lhyra;
+template<typename T, typename U, unsigned int SIZE> class Lhyra;
 
-template<typename T, typename U>
+/*
+When subclassing Optimizer, Constructor should be basically blank.
+Instead, use init(), which is called by Lhyra, to do initializaiton.
+*/
+template<typename T, typename U, unsigned int SIZE>
 class Optimizer {
 protected:
-    Lhyra<T, U> * lhyra;
-    double testy;
+    Lhyra<T, U, SIZE> * lhyra;
 public:
-    Optimizer(Lhyra<T, U> * l): lhyra(l) { testy = 0; }
-    virtual Solver<T, U> & solver(const std::vector<double> & features) = 0;
+    Optimizer() { lhyra = NULL; }
+    void set_lhyra(Lhyra<T, U, SIZE> * l) {
+        lhyra = l;
+        init();
+    }
+    virtual void init() = 0;
+    virtual Solver<T, U> & solver(const std::array<double, SIZE> & features) = 0;
 };
 
-template<typename T, typename U>
+// For debugging:
+#include <iostream>
+// End
+
+template<typename T, typename U, unsigned int SIZE>
 class Lhyra {
 private:
-    std::vector<Solver<T, U>> solvers;
-    DataStore<T> * datastore;
-    FeatureExtractor<T> * extractor;
-    Optimizer<T, U> * optimizer;
-    bool vocal;
+    FeatureExtractor<T, SIZE> * extractor;
+    Optimizer<T, U, SIZE> * optimizer;
 public:
+    // The optimizer needs to see these:
+    DataStore<T> * datastore;
+    std::vector<Solver<T, U>> solvers;
+    bool vocal;
     std::vector<double> times;
     
-    Lhyra(const std::vector<Solver<T, U>> & s,
-              const DataStore<T> * ds,
-              const FeatureExtractor<T> * e,
[…]

### 2019-05-13 c++/lhyra_test.cpp

@@ -4,19 +4,6 @@
 #include <stdlib.h>
 #include <time.h>
 #include <iostream>
-
-/* FOR TESTING LHYRA:
-
-Order of initialization:
-
-1. Create datastore
-2. Create feature extractor
-3. Create optimizer<in_type, out_type, extractor.size>
-4. Create LHYRA
-
-
-*/
-
 constexpr int LIST_SIZE = 4;
 
 struct random_list {

### 2019-05-13 c++/linopt.hpp

@@ -57,18 +57,16 @@ public:
 
         return result;
     }
-};
 
-template<typename T>
-class Test {
-protected:
-public:
-    double testd;
-    Test(double d) { testd = d; }
+    void pretty_print(std::ostream & o) {
+        o << "[Coeffs=<";
+        for(int i = 0; i < SIZE; i++) o << coeffs[i] << ',';
+        o << ">; Bias=" << bias << ']';
+    }
 };
 
 template<typename T, typename U, unsigned int SIZE>
-class LinOptimizer: public Optimizer<T, U>, public Test<T> {
+class LinOptimizer: public Optimizer<T, U, SIZE> {
 private:
     double max_eps, min_eps, eps;
     double totaltime;
@@ -77,18 +75,17 @@ private:
     bool training;
     unsigned int num_solvers;
     
-    std::vector<LinearRegression<SIZE>> regr;
-    
 public:
-    LinOptimizer(Lhyra<T, U> * l, double maxeps = 0.9, double mineps = 0.1) :
-                Optimizer<T, U>(l), Test<T>(maxeps) {
-        
-        testd += 1;
+    std::vector<LinearRegression<SIZE>> regr; // Exposing temporarily to get coeffs.
+    
+    LinOptimizer(): Optimizer<T, U, SIZE>() {
         max_eps = 0.9;
         min_eps = 0.1;
-        eps = max_eps;
+    }
+
+    void init() {
         
-        num_solvers = l->solvers.size();        
+        num_solvers = this->lhyra->solvers.size();        
         regr = std::vector< LinearRegression<SIZE> >();
         regr.reserve(num_solvers);
         for(int i = 0; i < num_solvers; i++) {
@@ -97,6 +94,8 @@ public:
         
         training = false;
         totaltime = 0;
+
+        eps = max_eps;
     }
     
     void train(int iters, int sample, bool log) {
@@ -107,25 +106,28 @@ public:
         auto features = std::vector< std::vector< std::array<double, SIZE> > >(num_solvers);
         auto times = std::vector< std::vector<double> >(num_solvers);
         
+        //std::cout << "Checkpoint 1.2" << std::endl;
         for(int episode = 0; episode < iters; episode++) {
             totaltimes.push_back(0);
             
             eps = max_eps + (episode/(iters-1))*(min_eps-max_eps);
             
-            auto data = lhyra->datastore->get_data(sample);
+            auto data = this->lhyra->datastore->get_data(sample);
[…]

### 2019-05-13 c++/make_graph.py

@@ -1,0 +1,24 @@
+from matplotlib import pyplot as plt
+from math import log2
+
+xs = []
+lhyra = []
+merge = []
+insertion = []
+quick = []
+
+with open('data.txt', 'r') as f:
+	for line in f.readlines():
+		x, l, m, i, q = tuple(line.split(' '))
+		xs.append(log2(int(x)))
+		lhyra.append((int(l)) / int(l))
+		merge.append((int(m)) / int(l))
+		insertion.append((int(i))/ int(l))
+		quick.append((int(q))/ int(l))
+
+plt.plot(xs, lhyra, label='lhyra')
+plt.plot(xs, merge, label='merge')
+plt.plot(xs, insertion, label='insertion')
+plt.plot(xs, quick, label='quick')
+plt.legend()
+plt.show()

### 2019-05-13 c++/plot_progress.py

@@ -1,0 +1,9 @@
+from matplotlib import pyplot as plt
+
+with open('times.txt', 'r') as f:
+	s = f.read()
+
+times = [float(t) / 10**6 for t in s.split('\n') if len(t) > 0]
+
+plt.plot(list(range(len(times))), times)
+plt.show()

### 2019-05-13 c++/sort_test.cpp

@@ -1,0 +1,195 @@
+#include "lhyra.hpp"
+#include "linopt.hpp"
+#include "sorting.hpp"
+
+#include <vector>
+#include <stdlib.h>
+#include <iostream>
+#include <fstream>
+
+
+// Data generator for sorting. Yes, you modify the globals to change the behavior. Sorry.
+int LIST_LENGTH = 1000;
+double NEARLY_SORTED = 0;
+void shuffle_slice(std::vector<double> & a, int start, int stop) {
+	int i = start;
+	while(i < stop-1) {
+		int idx = rand() % (1+stop-i) + i;
+		double temp = a[i];
+		a[i] = a[idx];
+		a[idx] = temp;
+		i++;
+	}
+}
+std::vector<double> random_list() {
+	std::vector<double> list;
+	list.reserve(LIST_LENGTH);
+	for(int i = 0; i < LIST_LENGTH; i++) {
+		list.emplace_back(i);
+	}
+	if(double(rand()) / RAND_MAX < NEARLY_SORTED) {
+		// Create 8 sections of length 5 which are shuffled:
+		for(int i = 0; i < 8; i++) {
+			int start = rand() % (LIST_LENGTH-5);
+			shuffle_slice(list, start, start+5);
+		}
+	}
+	else shuffle_slice(list, 0, LIST_LENGTH-1);
+	return list;
+}
+
+// Feature extractor for sorting
+constexpr int FEATURE_SIZE = 4;
+class SortFeatureExtractor: public FeatureExtractor<std::vector<double>, FEATURE_SIZE> {
+public:
+	SortFeatureExtractor(){}
+
+	double random_sorted(const std::vector<double> & data, int num_pts) {
+		std::vector<int> pts;
+		pts.reserve(num_pts);
+		for(int i = 0; i < num_pts; i++) pts.emplace_back(rand() % data.size());
+		// Insertion sort pts
+		for(int i = 0; i < num_pts-1; i++) for(int j = i; j >= 0 && pts[j+1] < pts[j]; j--) {
+			int temp = pts[j];
+			pts[j] = pts[j+1];
+			pts[j+1] = temp;
+		}
+		// Confirm that samples are in order.
+		for(int i = 1; i < num_pts; i++) {
+			//std::cout << pts[i-1] << ' ' << pts[i] << "        ";
+			//std::cout.flush();
[…]

### 2019-05-13 c++/sorting.cpp

@@ -1,71 +1,0 @@
-#include <vector>
-#include <iostream>
-
-std::vector<double> merge_sort(std::vector<double> data) {
-	if (data.size() <= 1) return data;
-
-	int halfway = data.size()/2;
-
-	auto sorted_first = merge_sort(std::vector<double>(data.begin(), data.begin() + halfway));
-	auto sorted_second = merge_sort(std::vector<double>(data.begin()+halfway, data.begin() + data.size()));
-
-	auto joined = std::vector<double>();
-
-	int j=0, i=0;
-	while (i+j < data.size()) {
-		if (i == sorted_first.size() || (j != sorted_second.size() and sorted_first[i] > sorted_second[j]))
-			joined.push_back(sorted_second[j++]);
-		else joined.push_back(sorted_first[i++]);
-	}
-
-	return joined;
-}
-
-std::vector<double> insertion_sort(std::vector<double> data) {
-	if (data.size() <= 1) return data;
-
-	std::vector<double> sorted = data;
-
-	for (int i=0; i<data.size()-1; i++) {
-		int j = i;
-		while (j >= 0 && sorted[j+1] < sorted[j]) {
-			double temp = sorted[j];
-			sorted[j] = sorted[j+1];
-			sorted[j+1] = temp;
-			j--;
-		}
-	}
-
-	return sorted;
-}
-
-std::vector<double> quick_sort(std::vector<double> data) {
-	if (data.size() <= 1) return data;
-
-	std::vector<double> sorted = data;
-
-	double pivot = sorted[sorted.size()/2];
-	int i=0, j=sorted.size()-1;
-
-	int pi;
-	while (true) {
-		while (sorted[i] < pivot) i++;
-		while (sorted[j] > pivot) j--;
-		if (i >= j) {
-			pi = j;
-			break;
-		}
-
-		double temp = sorted[i];
-		sorted[i] = sorted[j];
-		sorted[j] = temp;
-	}
-
-	auto sorted_first = quick_sort(std::vector<double>(sorted.begin(), sorted.begin() + pi));
-	auto sorted_second = quick_sort(std::vector<double>(sorted.begin()+pi+1, sorted.begin() + sorted.size()));
-
-	sorted_first.push_back(sorted[pi]);
-	sorted_first.insert(sorted_first.end(), sorted_second.begin(), sorted_second.end());
-
-	return sorted_first;
-}

### 2019-05-13 c++/sorting.hpp

@@ -1,0 +1,85 @@
+#pragma once
+
+#include <vector>
+#include <iostream>
+#include <functional>
+
+std::vector<double> merge_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+	std::vector<double> data{data2};
+	if (data.size() <= 1) return data;
+
+	int halfway = data.size()/2;
+
+	auto sorted_first = hook(std::vector<double>(data.begin(), data.begin() + halfway));
+	auto sorted_second = hook(std::vector<double>(data.begin()+halfway, data.begin() + data.size()));
+
+	auto joined = std::vector<double>();
+
+	int j=0, i=0;
+	while (i+j < data.size()) {
+		if (i == sorted_first.size() || (j != sorted_second.size() and sorted_first[i] > sorted_second[j]))
+			joined.push_back(sorted_second[j++]);
+		else joined.push_back(sorted_first[i++]);
+	}
+
+	return joined;
+}
+std::vector<double> merge_handle(const std::vector<double> & data) {
+	std::function<std::vector<double>(const std::vector<double> &)> mh = merge_handle;
+	return merge_sort(data, mh);
+}
+
+std::vector<double> insertion_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+	std::vector<double> data{data2};
+	if (data.size() <= 1) return data;
+
+	std::vector<double> sorted = data;
+
+	for (int i=0; i<data.size()-1; i++) {
+		int j = i;
+		while (j >= 0 && sorted[j+1] < sorted[j]) {
+			double temp = sorted[j];
+			sorted[j] = sorted[j+1];
+			sorted[j+1] = temp;
+			j--;
+		}
+	}
+
+	return sorted;
+}
+
+std::vector<double> quick_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+	std::vector<double> data{data2};
+	if (data.size() <= 1) return data;
+
+	std::vector<double> sorted = data;
+
+	double pivot = sorted[sorted.size()/2];
+	int i=0, j=sorted.size()-1;
+
+	int pi;
+	while (true) {
+		while (sorted[i] < pivot) i++;
+		while (sorted[j] > pivot) j--;
+		if (i >= j) {
+			pi = j;
+			break;
+		}
+
+		double temp = sorted[i];
[…]

### 2019-05-16 c++/lhyra.hpp
bugfix and dockerfile and debugging

@@ -103,20 +103,26 @@ public:
     }
     U operator()(const T & data) {
         
+
         auto t1 = std::chrono::high_resolution_clock::now();
         
+        int time_slot = times.size();
+        times.push_back(0);
+        std::cout << "size: " << times.size() << std::endl;
+
         //std::cout << "Checkpoint 5.1" << std::endl;
         auto features = (*extractor)(data);
+        
         //std::cout << "Checkpoint 5.2" << std::endl;
+        //std::cout << "me: " << this << std::endl;
         auto solver = optimizer->solver(features);
         //std::cout << "Checkpoint 5.3" << std::endl;
-        std::function<U(const T&)> hook = std::bind(*this, std::placeholders::_1);
+        std::function<U(const T&)> hook = *this;
         auto sol = solver(data, hook);
         
         auto t2 = std::chrono::high_resolution_clock::now();
         
-        times.push_back(std::chrono::duration_cast<std::chrono::nanoseconds>(t2 - t1).count());
-        //std::cout << times.back() << std::endl;
+        times[time_slot] = std::chrono::duration_cast<std::chrono::nanoseconds>(t2 - t1).count();
         
         return sol;
     }

### 2019-05-16 c++/dummyopt.hpp
fixed major buggos

@@ -1,0 +1,13 @@
+#pragma once
+
+#include "lhyra.hpp"
+
+template<typename T, typename U, unsigned int SIZE>
+class DummyOptimizer: public Optimizer<T, U, SIZE> {
+public:
+	DummyOptimizer(): Optimizer<T, U, SIZE>() {}
+	void init() {}
+	Solver<T, U, SIZE> & solver(const std::array<double, SIZE> & features) {
+        return this->lhyra->solvers[0];
+	}
+};

### 2019-05-16 c++/lhyra.hpp

@@ -13,9 +13,9 @@ public:
 template<typename T>
 class DataGenerator: public DataStore<T> {
 private:
-    std::function<T()> method;
+    T(*method)();
 public:    
-    DataGenerator(const std::function<T()> & m) {
+    DataGenerator(T(*m)()) {
         method = m;
     }
     std::vector<T> get_data(int batch_size) {
@@ -39,17 +39,19 @@ public:
     virtual U operator()(const T & t) = 0;
 }*/
 
-template<typename T, typename U>
-using Solver = std::function<U(const T &, std::function<U(const T &)> &)>;
+
+// Early declaration for the compiler
+template<typename T, typename U, unsigned int SIZE> class Lhyra;
+
+
+template<typename T, typename U, unsigned int SIZE>
+using Solver = U(*)(const T &, Lhyra<T, U, SIZE> &);
 
 template<typename T, unsigned int SIZE>
 class FeatureExtractor {
 public:
     virtual std::array<double, SIZE> operator()(const T & t) = 0;
 };
-
-// Early declaration for the compiler
-template<typename T, typename U, unsigned int SIZE> class Lhyra;
 
 /*
 When subclassing Optimizer, Constructor should be basically blank.
@@ -66,7 +68,7 @@ public:
         init();
     }
     virtual void init() = 0;
-    virtual Solver<T, U> & solver(const std::array<double, SIZE> & features) = 0;
+    virtual Solver<T, U, SIZE> & solver(const std::array<double, SIZE> & features) = 0;
 };
 
 // For debugging:
@@ -81,11 +83,11 @@ private:
 public:
     // The optimizer needs to see these:
     DataStore<T> * datastore;
-    std::vector<Solver<T, U>> solvers;
+    std::vector<Solver<T, U, SIZE>> solvers;
     bool vocal;
     std::vector<double> times;
     
-    Lhyra(std::vector<Solver<T, U>> & s,
+    Lhyra(std::vector<Solver<T, U, SIZE>> & s,
               DataStore<T> * ds,
               FeatureExtractor<T, SIZE> * e,
               Optimizer<T, U, SIZE> * o):
@@ -103,6 +105,11 @@ public:
     }
     U operator()(const T & data) {
         
+        size_t index = times.size();
+        times.emplace_back(0);
+
+        // std::cout << this << std::endl;
+
         auto t1 = std::chrono::high_resolution_clock::now();
         
         //std::cout << "Checkpoint 5.1" << std::endl;
@@ -110,12 +117,11 @@ public:
[…]

### 2019-05-16 c++/linopt.hpp

@@ -120,6 +120,8 @@ public:
                 
                 totaltimes.back() += this->lhyra->times[0];
                 
+                // std::cout << epoch_choices.size() << ' ' << this->lhyra->times.size() << std::endl;
+
                 for(int i = 0; i < epoch_choices.size(); i++) {
                     features[epoch_choices[i]].push_back(epoch_features[i]);
                     times[epoch_choices[i]].push_back(this->lhyra->times[i]);
@@ -157,7 +159,7 @@ public:
         train(100, 40, true);
     }
     
-    Solver<T, U> & solver(const std::array<double, SIZE> & features) {
+    Solver<T, U, SIZE> & solver(const std::array<double, SIZE> & features) {
         int action;
         if(training && double(rand()) / RAND_MAX < eps) {
             action = rand() % num_solvers;

### 2019-05-16 c++/make_graph.py

@@ -6,19 +6,22 @@ lhyra = []
 merge = []
 insertion = []
 quick = []
+std = []
 
 with open('data.txt', 'r') as f:
 	for line in f.readlines():
-		x, l, m, i, q = tuple(line.split(' '))
+		x, l, m, i, q, s = tuple(line.split(' '))
 		xs.append(log2(int(x)))
-		lhyra.append((int(l)) / int(l))
-		merge.append((int(m)) / int(l))
-		insertion.append((int(i))/ int(l))
-		quick.append((int(q))/ int(l))
+		lhyra.append(log2(int(l)))
+		merge.append(log2(int(m)))
+		insertion.append(log2(int(i)))
+		quick.append(log2(int(q)))
+		std.append(log2(int(s)))
 
 plt.plot(xs, lhyra, label='lhyra')
 plt.plot(xs, merge, label='merge')
 plt.plot(xs, insertion, label='insertion')
 plt.plot(xs, quick, label='quick')
+plt.plot(xs, std, label='std')
 plt.legend()
 plt.show()

### 2019-05-16 c++/sort_test.cpp

@@ -1,11 +1,13 @@
 #include "lhyra.hpp"
 #include "linopt.hpp"
+#include "dummyopt.hpp"
 #include "sorting.hpp"
 
 #include <vector>
 #include <stdlib.h>
 #include <iostream>
 #include <fstream>
+#include <algorithm>
 
 
 // Data generator for sorting. Yes, you modify the globals to change the behavior. Sorry.
@@ -39,7 +41,6 @@ std::vector<double> random_list() {
 }
 
 // Feature extractor for sorting
-constexpr int FEATURE_SIZE = 4;
 class SortFeatureExtractor: public FeatureExtractor<std::vector<double>, FEATURE_SIZE> {
 public:
 	SortFeatureExtractor(){}
@@ -87,16 +88,6 @@ Order of initialization:
 
 */
 
-std::vector<double> foo(std::vector<double> i,
-		std::function<std::vector<double>(std::vector<double>)> hook) {
-	if(i.empty()) return std::vector<double>{};
-	double back = i.back();
-	i.pop_back();
-	std::vector<double> answer = hook(i);
-	answer.push_back(back);
-	return answer;
-}
-
 
 int main() {
 
@@ -108,7 +99,7 @@ int main() {
 
 	// Create vector of solvers:
 
-	std::vector<Solver<std::vector<double>, std::vector<double>>> solvers;
+	std::vector<Solver<std::vector<double>, std::vector<double>, FEATURE_SIZE>> solvers;
 	solvers.push_back(&merge_sort);
 	solvers.push_back(&insertion_sort);
 	solvers.push_back(&quick_sort);
@@ -138,7 +129,7 @@ int main() {
 
 	//std::cout << "Checkpoint 1" << std::endl;
 
-	linopt.train();
+	linopt.train(200, 40, true);
 
 	//std::cout << "Checkpoint " << std::endl;
 
@@ -151,6 +142,38 @@ int main() {
 
 	std::ofstream f("data.txt");
 
+	// Create Lhyras for merge, insertion, and quick sort.
+	std::vector<Solver<std::vector<double>, std::vector<double>, FEATURE_SIZE>> merge_solvers;
+	std::vector<Solver<std::vector<double>, std::vector<double>, FEATURE_SIZE>> insert_solvers;
+	std::vector<Solver<std::vector<double>, std::vector<double>, FEATURE_SIZE>> quick_solvers;
[…]

### 2019-05-16 c++/sorting.hpp

@@ -1,17 +1,21 @@
 #pragma once
 
+#include "lhyra.hpp"
+
 #include <vector>
 #include <iostream>
-#include <functional>
 
-std::vector<double> merge_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+constexpr int FEATURE_SIZE = 4;
+
+std::vector<double> merge_sort(const std::vector<double> & data2,
+		Lhyra<std::vector<double>, std::vector<double>, FEATURE_SIZE> & lhyra) {
 	std::vector<double> data{data2};
 	if (data.size() <= 1) return data;
 
 	int halfway = data.size()/2;
 
-	auto sorted_first = hook(std::vector<double>(data.begin(), data.begin() + halfway));
-	auto sorted_second = hook(std::vector<double>(data.begin()+halfway, data.begin() + data.size()));
+	auto sorted_first = lhyra(std::vector<double>(data.begin(), data.begin() + halfway));
+	auto sorted_second = lhyra(std::vector<double>(data.begin()+halfway, data.begin() + data.size()));
 
 	auto joined = std::vector<double>();
 
@@ -24,12 +28,8 @@ std::vector<double> merge_sort(const std::vector<double> & data2, std::function<
 
 	return joined;
 }
-std::vector<double> merge_handle(const std::vector<double> & data) {
-	std::function<std::vector<double>(const std::vector<double> &)> mh = merge_handle;
-	return merge_sort(data, mh);
-}
-
-std::vector<double> insertion_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+std::vector<double> insertion_sort(const std::vector<double> & data2,
+		Lhyra<std::vector<double>, std::vector<double>, FEATURE_SIZE> & lhyra) {
 	std::vector<double> data{data2};
 	if (data.size() <= 1) return data;
 
@@ -48,7 +48,8 @@ std::vector<double> insertion_sort(const std::vector<double> & data2, std::funct
 	return sorted;
 }
 
-std::vector<double> quick_sort(const std::vector<double> & data2, std::function<std::vector<double>(const std::vector<double> &)> & hook) {
+std::vector<double> quick_sort(const std::vector<double> & data2,
+		Lhyra<std::vector<double>, std::vector<double>, FEATURE_SIZE> & lhyra) {
 	std::vector<double> data{data2};
[…]

### 2019-05-16 c++/make_graph.py
travis

@@ -24,4 +24,5 @@ plt.plot(xs, insertion, label='insertion')
 plt.plot(xs, quick, label='quick')
 plt.plot(xs, std, label='std')
 plt.legend()
-plt.show()
+
+plt.savefig('relative_plot.png')

### 2019-05-16 c++/plot_progress.py

@@ -6,4 +6,5 @@ with open('times.txt', 'r') as f:
 times = [float(t) / 10**6 for t in s.split('\n') if len(t) > 0]
 
 plt.plot(list(range(len(times))), times)
-plt.show()
+
+plt.savefig('progress.png')

### 2019-05-16 python/imgur.py

@@ -1,0 +1,12 @@
+from imgurpython import ImgurClient
+
+CLIENT_ID = '3fa744e80ad2ae7'
+CLIENT_SECRET = '59ba07338b6f4c79d0a9701134051a2bffdb2c95'
+
+images = ["../c++/progress.png", "../c++/relative_plot.png"]
+
+client = ImgurClient(CLIENT_ID, CLIENT_SECRET)
+
+for image in images:
+	response = client.upload_from_path(image, anon=True)
+	print(response["link"])

### 2019-05-16 c++/tests.sh
forgot to commit scriptrunner

@@ -1,0 +1,17 @@
+echo "Running code..."
+
+./sort_test
+
+echo "Generating training plot..."
+
+python plot_progress.py
+
+echo "Generating performance plot..."
+
+python make_graph.py
+
+echo "Uploading to imgur..."
+
+cd ../python
+
+python imgur.py

### 2019-05-16 c++/lhyra.hpp
too much printing.

@@ -112,7 +112,6 @@ public:
         
         int time_slot = times.size();
         times.push_back(0);
-        std::cout << "size: " << times.size() << std::endl;
 
         //std::cout << "Checkpoint 5.1" << std::endl;
         auto features = (*extractor)(data);

### 2019-05-16 c++/tests.sh
clean up

@@ -1,3 +1,7 @@
+echo "Compiling and linking..."
+
+g++ -I/usr/include -c sort_test.cpp -std=c++14 -o sort_test.o && g++ -L/usr/lib sort_test.o -lgsl -lgslcblas -lm -o sort_test -std=c++14
+
 echo "Running code..."
 
 ./sort_test

### 2019-05-16 paper/main.tex
paper draft without figures

\documentclass{article}
\usepackage{hyperref, graphicx, floatrow}

\usepackage[letterpaper, margin=1in]{geometry}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0pt}

\title{Lhyra: Learned HYbrid Recursive Algorithms}

\author{
    Josh Gruenstein\\\texttt{jgru@mit.edu} \and Lior Hirschfeld\\\texttt{liorh@mit.edu} \and
    Benjamin Spector\\\texttt{spectorb@mit.edu}
}

\date{6.890 Spring 2019 Final Project}

\begin{document}

\maketitle

\section{Introduction}
A large number of computer science problems lend themselves naturally to recursive solutions. Of these, if more than one known algorithm exists, choosing between them may be difficult. Even if asymptotic, worst-case run time is known, this may have no bearing on real world use cases, which oftentimes operate on small inputs or data that is not representative of the entire space. Generally, a decision tree is chosen manually by an expert, who benchmarks the performance of each algorithm and selects whichever appears to perform most efficiently. A clear example of this is in the C++ STL, in which the standard algorithm is Introsort, which calls Quicksort until hitting a certain recursion depth, where it switches to Heapsort. This approach is far from ideal, not only because it is so time consuming, but also because decision making will be restricted only to those features which the expert tested most rigorously.

Instead, we propose that this process should be automated through the introduction of machine learning. For this project, we designed and implemented Lhyra, a framework designed to automatically find efficient recursive trees given arbitrary solvers, data distributions, and optimization criteria. Lhyra was largely inspired by Professor Kraska's work on learned B-Trees and the Recursive Model Index \cite{kraska}, as we believed that, through high-level abstractions, it might be possible to apply a similar methodology to a large number of recursive data structures and algorithms. Below, we discuss Lhyra's components in more detail and present our results when applying Lhyra to two common problems, \textsc{Closest Pair} and \textsc{Sorting}.

\section{Framework}
A Lhyra instance is described by the following components:

\begin{itemize}
	\item \textbf{Bag of Solvers $S$.} Each solver $s_i \in S$ is equipped to solve an instance of the problem at hand, potentially by calling the Lhyra instance on sub-problems. Each solver $s_i$ may also have a set of hyper-parameters $h_{ij}$.

	\item \textbf{Data Generator / Training Data Set $D$.} In the training phase the problems in this set / generated by this generator are used by the optimizer to learn how to select a solver $s_i$ from S given problem features.

	\item \textbf{Feature Extractor $F(x)$.} Our goal is to learn how to select the ideal solver based on a given input, thus we need a nice way of extracting useful features about an input. For example, a feature extractor for a sorting Lhyra instance may return a vector containing the length of the list, number of unique elements, and how sorted it already is. Defaults to a vector of length 1 containing the depth of the recursion.

	\item \textbf{Optimizer $O$.} Given a feature extractor, a training data set, and a bag of solvers, the optimizer learns how to select a solver given a vector of problem features.  Optimizers should default to minimizing computation time and/or memory, but can take an arbitrary cost function $C(x,y)$ that produces some cost to be minimized given a problem and solution. 
\end{itemize}

Once initialized, a call to a Lhyra instance solves a problem $x$ through the following steps:

\begin{enumerate}
	\item Compute $F(x)$.
	\item Pass $F(x)$ into the solver selector learned by the optimizer, which returns a parametrized solver $s_{ih}$.
	\item Compute $s_{ih}(x)$.
	\begin{enumerate}
		\item For each sub-problem $y_i$ identified by $s_{ih}$, repeat the above process.
		\item Combine the sub-problems and return a result.
	\end{enumerate}
\end{enumerate}

An ideal solver selector and feature extractor would be fast enough for the benefits gained by ideal solver selection to outweigh the additional overhead. After implementing Lhyra in both Python and C++, we found relative performance to be very impacted by language choice and implementation specifics.

\subsection{Optimizers}

The optimizer carries the responsibility of selecting the ideal solver, and training the model to perform said selection process.  An ideal Lhyra optimizer would satisfy the following conditions:

\begin{itemize}
    \item \textbf{Applicability.} The optimizer must learn using data that accessible. For example, gradient descent may be impossible if the relationship between input and cost is unclear.
    \item \textbf{Flexibility} Given a feature vector, the optimizer must be able to apply a sufficiently complex matching to extract meaningful data for solver selection.
    \item \textbf{Speed.} The optimizer must predict quickly enough that its overhead does not outweigh gains from selecting the proper solver. This component can be ignored if time is not an optimization criteria.
\end{itemize}

Although each optimizer we experimented with used a different training process, we decided to use a common overall framework in generating feedback times (our optimizer cost function $C$).  We chose to measure the entire time to execute a sub-problem, including recursive calls to (potentially different) solvers selected by Lhyra.  At face value, this is an odd choice compared to the far more straightforward method of simply measuring a solver's own execution time, not including that of its children.  It introduces strange cyclical dependencies, where Lhyra's choice of solver must depend on an ingrained understanding of its own decision making.  This also forces Lhyra's ``training'' process to be online, as timing measurements are only useful using the most up-to-date model parameters.

However, this is exactly the property we would want in an ideal optimizer.  Consider a solver which did minimal work, instead deciding to punt nearly all of the computation to its children.  A naive optimizer would unduly favor such a solver, and fail to recognize useful sequential relationships in solvers: for example, that solver B might be best following solver A.  Although this methodology is harder to train, it is ideal presuming the model is able to converge at all (which we empirically do with some optimizers).

\subsubsection{Neural Network Classifiers}

A natural optimizer architecture might feed our feature vector through a neural network, which outputs a one-hot vector representing solver selection. As we can make this neural network arbitrary large, this system is guaranteed to satisfy our condition of flexibility. However, it suffers in other respects.  As our classifier gets larger, classification latency would suffer.  Techniques designed to speed up neural network execution (such as GPUs) largely rely on batching for speed and are do not significantly improve latency for single inferences, limiting the size of potential networks.

Training is also a challenge. Gradient descent is incompatible with this architecture, because there exists no discernible relationship between network weights and cost: without exhausting all available options, it is impossible to determine whether the ideal solver was selected.  We attempted to train a single perceptron classifier using REINFORCE \cite{reinforce}, a member of the policy gradients family that uses reward signals to estimate a classifier gradient.  However, we found our classifier was unable to converge on reasonable solutions, as the estimated gradients were far too noisy.

We were able to obtain somewhat reasonable results training the same model from above using naive hill-climbing, randomly perturbing weights and keeping perturbations that led to performance improvements.  However, our results were outperformed by better optimizers (such as our linear regressor), and we do not think this training method would scale to different and more complex problems due to its fragility.

\subsubsection{Linear Regressors}

An alternative architecture trains a set of linear regressors, where each predicts execution costs for a call-tree starting at a certain solver type.  For any given feature vector, the optimizer predicts with each regressor and selects the solver corresponding to the regressor with smallest output. Each regressor can be easily trained through linear least-squares.

The primary concern with this method is that it lacks flexibility. With only a single perceptron, a linear regressor lacks the ability to extract nonlinear relationships between features. For both \textsc{Closest Pair} and \textsc{Sorting}, we did not find this was not problematic, but this deficiency should be considered when solver selection is more complex. One could imagine remedying this issue either with automatic feature extraction or a more powerful (non-linear and non-convex) regressor.

Even with a convergent optimizer, we struggled with overhead from running our regressors.  For example, in order to make the overhead on recursive calls in our Python implementation net-positive, we had to perform micro-optimizations such as swapping out Scikit-learn's \texttt{LinearRegression} \cite{sklearn} class's inference methods for our own.

\section{Experiments}

In order to develop and test Lhyra, we built implementations in both Python and C++.  We will  show here results from C++ benchmarks, as we found relative performance measurement to be difficult in Python due to overhead such as garbage collection.

\subsection{Closest Pair of Points}

Consider the \textsc{Closest Pair} problem, in which given a set of $n$ points $P$ you must identify a pair of unique points $(p_i,p_j)$ to minimize distance $||p_i - p_j||_2$.  This is a common problem, but not quite common enough for hyper-optimized hand-coded adaptive implementations to exist.  However, there are two common algorithms for solving this problem.

\begin{enumerate}
    \item \textbf{Brute force.}  In $O(n^2)$ time, exhaustively enumerate all possible pairs of points, and choose the one with the smallest distance.
    \item \textbf{Divide \& conquer.}  Partition the points by the median $x$ coordinate, and recurse on the points to the left and right of the dividing line, $P_l$ and $P_r$ respectively.  Then, for each point within $\min(d(P_r),d(P_r))$ of the dividing line, enumerate all possible pairs to see if one has a shorter distance than $d(P_r)$ or $d(P_l)$, where $d$ is the smallest pair distance in that set.  It can be shown geometrically that the number of pairs within this distance of the dividing line is $O(n)$ \cite{clrs}, so the algorithm overall is $O(n\log n)$.
\end{enumerate}

We implemented a Lhyra instance with the above two solvers, using our linear regressor-based optimizer and extracting set size $n$ as our only feature.  With only seconds of training, Lhyra was able to substantially outperform both algorithms for nearly all $n$ by learning to use the divide-and-conquer method at high $n$, and the brute-force method at lower $n$.

\vspace{0.5cm}
\begin{figure}[!h]
    \centering
    \begin{floatrow}
        \ffigbox{
            \caption{\textsc{Closest Pair} training curve}
        }{
            \includegraphics[width=5.5cm]{example-image}
        }

        \ffigbox{
            \caption{\textsc{Closest Pair} relative performance}
        }{%
            \includegraphics[width=5.5cm]{example-image}
        }
    \end{floatrow}
\end{figure}

\subsection{List Sorting}

A more complex recursive problem is \textsc{Sorting}, producing an ordered list from a potentially unordered one.  We chose the following common sorting methods for our bag of solvers: Insertion Sort ($O(n^2)$), Merge Sort ($O(n \log n)$), and Quicksort (also $O(n \log n)$). Due to their ubiquity, we will not review how these algorithms work.

We implemented a Lhyra instance with the three solvers, the same optimizer as above, and two features: the list length $n$, and sortedness $n*s$, calculated in $O(1)$ time by randomly sampling 10 elements from the list and checking their orderedness $s$, then multiplying by $n$ to normalize to our other feature.  Below are results from this instance using purely-random lists, showing that Lhyra dramatically outperforms the fastest algorithm Quicksort at large $n$, and matches Insertion Sort at small $n$.

\vspace{0.5cm}
\begin{figure}[!ht]
    \centering
    \begin{floatrow}
        \ffigbox{
            \caption{Random \textsc{Sorting} training curve}
        }{
            \includegraphics[width=5.5cm]{example-image}
        }

        \ffigbox{
            \caption{Random \textsc{Sorting} relative performance}
        }{%
            \includegraphics[width=5.5cm]{example-image}
        }
    \end{floatrow}
\end{figure}

However, Lhyra can perform even better with interesting and changing data distributions.  Take the following example, where lists are already 50\% sorted.

\vspace{0.5cm}
\begin{figure}[!ht]
    \centering
    \begin{floatrow}
        \ffigbox{
            \caption{Partially-sorted \textsc{Sorting} training curve}
        }{
            \includegraphics[width=5.5cm]{example-image}
        }

        \ffigbox{
            \caption{Partially-sorted \textsc{Sorting} relative performance}
        }{%
            \includegraphics[width=5.5cm]{example-image}
        }
    \end{floatrow}
\end{figure}

Lhyra learns that our implementation of Insertion Sort is $O(n)$ on sorted lists, and dynamically exploits that knowledge to widen its performance gap against other solvers.

\section{Discussion}

% parametrizing solvers: random variables, etc
% learned feature extraction
One potential future optimization is pruning the \textit{Optimizer}. Shrinking the network's size would save time with each prediction, perhaps resulting in a significant speedup over many recursive calls. This process would also inform us of useless features. If any edges attached to the input are deleted from the network during pruning, then we can automatically remove the corresponding feature from the \textit{Feature Extractor}, resulting in additional performance gains.

A natural extension of Lhyra is introducing learned feature extraction. This process may involve combining the \textit{Feature Extractor} and \textit{Optimizer} components. Instead of passing in a feature vector to the \textit{Optimizer}, it receives raw data as input. For example, if operating on a graph, we could make use of Message Passage Neural Networks, which have recently been shown to dramatically improve performance over traditional features on tasks like molecular property prediction \cite{GilmerSRVD17}. Clearly, this process would require more computation, so this method would likely only be worthwhile on tasks where cost is lessened dramatically by good solver selection.

An application of Lhyra which we have not yet explored is online learning in production environments.  If the data distribution changes over time, Lhyra could continuously use evaluation data to train its solver picker.  This is actually doable for both \textsc{Sorting} and \textsc{Closest Pair}, as training for the whole dataset occurs in seconds, so miniature iterative updates could easily be done in production while remaining net-positive.

We would also like to investigate Lhyra's effectiveness at optimizing criteria other than time. One option would be to explore polynomial time approximation algorithms to NP-hard problems. In this scenario, cost could be a weighted function of evaluation time and accuracy, depending on the priorities of the user.

\bibliographystyle{unsrt}
\bibliography{ref}

\appendix
\section{Code and Reproducibility}

Our research codebase is available under the MIT license at the below Github repository.  Our code structure is described in the file \texttt{README.md}.

\begin{center}
\url{https://github.com/joshuagruenstein/lhyra}
\end{center}

In order to maximize reproducibility, we bundled our codebase into a Docker container, and set up Continuous Integration through Travis CI to auto-generate plots whenever we push to our Github repository.  Thus, we have a transparent, reproducible pipeline for obtaining results with Lhyra.  You can check out the figures generated by our most recent commit by going to the following link, and navigating to the \texttt{i.imgur.com} links at the bottom of the console output.

\begin{center}
\url{https://travis-ci.org/joshuagruenstein/lhyra}
\end{center}

\end{document}

### 2019-05-16 c++/linopt.hpp
everything fixed

@@ -110,7 +110,7 @@ public:
         for(int episode = 0; episode < iters; episode++) {
             totaltimes.push_back(0);
             
-            eps = max_eps + (episode/(iters-1))*(min_eps-max_eps);
+            eps = max_eps + (double(episode)/(iters-1))*(min_eps-max_eps);
             
             auto data = this->lhyra->datastore->get_data(sample);
             //std::cout << "Checkpoint 1.3" << std::endl;

### 2019-05-16 c++/points.cpp

@@ -6,26 +6,26 @@
 #include <limits>
 #include <stdlib.h>
 
-struct point {
+struct Point {
 	double x;
 	double y;
 };
 
-double point_dist(point p1, point p2) {
+double point_dist(Point p1, Point p2) {
 	return sqrt(pow(p1.x - p2.x,2) + pow(p1.y - p2.y,2));
 }
 
-double point_dist(std::pair<point, point> points) {
+double point_dist(std::pair<Point, Point> points) {
 	return point_dist(points.first, points.second);
 }
 
-std::pair<point, point> points_brute(std::vector<point> points) {
+std::pair<Point, Point> points_brute(std::vector<Point> points) {
 	double min_dist =  std::numeric_limits<double>::max();
-	std::pair<point, point> closest_pair;
+	std::pair<Point, Point> closest_pair;
 
 	for (int i=0; i<points.size(); i++) {
 		for (int j=i+1; j<points.size(); j++) {
-			auto pair = std::pair<point,point>(points[i],points[j]);
+			auto pair = std::pair<Point,Point>(points[i],points[j]);
 			if (point_dist(pair) < min_dist) {
 				min_dist = point_dist(pair);
 				closest_pair = pair;
@@ -36,23 +36,23 @@ std::pair<point, point> points_brute(std::vector<point> points) {
 	return closest_pair;
 }
 
-std::pair<point, point> points_smart(std::vector<point> points) {
+std::pair<Point, Point> points_smart(std::vector<Point> points) {
 	if (points.size() <= 3) return points_brute(points);
 
-	auto psorted = std::vector<point>(points);
+	auto psorted = std::vector<Point>(points);
 	std::sort(psorted.begin(), psorted.end(), [](const auto& l, const auto& r) {
 		return l.x < r.x;
 	});
 
 	int halfway = psorted.size()/2;
-	auto left = points_smart(std::vector<point>(psorted.begin(), psorted.begin()+halfway));
-	auto right = points_smart(std::vector<point>(psorted.begin()+halfway, psorted.begin()+psorted.size()));
+	auto left = points_smart(std::vector<Point>(psorted.begin(), psorted.begin()+halfway));
+	auto right = points_smart(std::vector<Point>(psorted.begin()+halfway, psorted.begin()+psorted.size()));
 	
 	double mid = psorted[halfway].x;
 	double dleft = point_dist(left), dright = point_dist(right);
[…]

### 2019-05-16 c++/lhyra.hpp
points

@@ -122,6 +122,7 @@ public:
         //std::cout << "Checkpoint 5.3" << std::endl;
 
         auto sol = solver(data, *this);
+        //std::cout << "Checkpoint 5.4" << std::endl;
         
         auto t2 = std::chrono::high_resolution_clock::now();
         

### 2019-05-16 c++/linopt.hpp

@@ -120,7 +120,7 @@ public:
                 
                 totaltimes.back() += this->lhyra->times[0];
                 
-                // std::cout << epoch_choices.size() << ' ' << this->lhyra->times.size() << std::endl;
+                //std::cout << epoch_choices.size() << ' ' << this->lhyra->times.size() << std::endl;
 
                 for(int i = 0; i < epoch_choices.size(); i++) {
                     features[epoch_choices[i]].push_back(epoch_features[i]);

### 2019-05-16 c++/make_graph_points.py

@@ -1,0 +1,22 @@
+from matplotlib import pyplot as plt
+from math import log2
+
+xs = []
+lhyra = []
+smart = []
+brute = []
+
+with open('data.txt', 'r') as f:
+	for line in f.readlines():
+		x, l, s, b = tuple(line.split(' '))
+		xs.append(log2(int(x)))
+		lhyra.append(log2(int(l)))
+		smart.append(log2(int(s)))
+		brute.append(log2(int(b)))
+
+plt.plot(xs, lhyra, label='lhyra')
+plt.plot(xs, smart, label='smart')
+plt.plot(xs, brute, label='brute')
+plt.legend()
+
+plt.savefig('relative_plot_points.png')

### 2019-05-16 c++/point_compile.sh

@@ -1,0 +1,2 @@
+g++ -I/home/sydriax/gsl/include -c points.cpp -std=c++14
+g++ -L/home/sydriax/gsl/lib points.o -lgsl -lgslcblas -lm -o points -std=c++14

### 2019-05-16 c++/points.cpp

@@ -1,10 +1,13 @@
-#pragma once
-
 #include <vector>
 #include <iostream>
 #include <cmath>
 #include <limits>
 #include <stdlib.h>
+#include <algorithm>
+
+#include "lhyra.hpp"
+#include "linopt.hpp"
+#include "dummyopt.hpp"
 
 struct Point {
 	double x;
@@ -19,11 +22,14 @@ double point_dist(std::pair<Point, Point> points) {
 	return point_dist(points.first, points.second);
 }
 
-std::pair<Point, Point> points_brute(std::vector<Point> points) {
+std::pair<Point, Point> points_brute(const std::vector<Point> & points,
+	Lhyra<std::vector<Point>, std::pair<Point, Point>, 1> & lhyra) {
+	//std::cout << "Brute solver; size=" << points.size() << std::endl;
 	double min_dist =  std::numeric_limits<double>::max();
 	std::pair<Point, Point> closest_pair;
 
 	for (int i=0; i<points.size(); i++) {
+		//std::cout << "Boop" << std::endl;
 		for (int j=i+1; j<points.size(); j++) {
 			auto pair = std::pair<Point,Point>(points[i],points[j]);
 			if (point_dist(pair) < min_dist) {
@@ -32,12 +38,16 @@ std::pair<Point, Point> points_brute(std::vector<Point> points) {
 			}
 		}
 	}
+	//std::cout << "Leaving brute solver" << std::endl;
 
 	return closest_pair;
 }
 
-std::pair<Point, Point> points_smart(std::vector<Point> points) {
-	if (points.size() <= 3) return points_brute(points);
+std::pair<Point, Point> points_smart(const std::vector<Point> & points,
+	Lhyra<std::vector<Point>, std::pair<Point, Point>, 1> & lhyra) {
+	// std::cout << "Smart solver" << std::endl;
+	if (points.size() <= 3) return points_brute(points, lhyra);
+
 
 	auto psorted = std::vector<Point>(points);
 	std::sort(psorted.begin(), psorted.end(), [](const auto& l, const auto& r) {
@@ -45,8 +55,8 @@ std::pair<Point, Point> points_smart(std::vector<Point> points) {
 	});
 
 	int halfway = psorted.size()/2;
-	auto left = points_smart(std::vector<Point>(psorted.begin(), psorted.begin()+halfway));
[…]

### 2019-05-16 c++/make_graph.py
Updated graphing files.

@@ -23,6 +23,10 @@ plt.plot(xs, merge, label='merge')
 plt.plot(xs, insertion, label='insertion')
 plt.plot(xs, quick, label='quick')
 plt.plot(xs, std, label='std')
+
+plt.xlabel('Log of List Length')
+plt.ylabel('Log of Time (in microseconds)')
+
 plt.legend()
 
 plt.savefig('relative_plot.png')

### 2019-05-16 c++/make_graph_points.py

@@ -15,8 +15,12 @@ with open('data.txt', 'r') as f:
 		brute.append(log2(int(b)))
 
 plt.plot(xs, lhyra, label='lhyra')
-plt.plot(xs, smart, label='smart')
-plt.plot(xs, brute, label='brute')
+plt.plot(xs, smart, label='divide and conquer')
+plt.plot(xs, brute, label='brute force')
+
+plt.xlabel('Log of Number of Points')
+plt.ylabel('Log of Time (in microseconds)')
+
 plt.legend()
 
 plt.savefig('relative_plot_points.png')

### 2019-05-16 proposal/proposal.tex

@@
 A clear example of this is in the C++ STL, in which the standard algorithm is Introsort, which calls Quicksort until hitting a certain recursion depth, where it switches to Heapsort.
-We observe that it would be better that these recursive structures would adapt to their distributions and objective functions, so that not only could they lessen the workload for the expert, but also better explore the potential synergies of the possible algorithms.
+We observe that it would be better if these recursive structures adapted to their distributions and objective functions, so that not only could they lessen the workload for the expert, but also better explore potential algorithmic `synergies'.
 
@@
 Regardless of the results, our expansive benchmarking will allow us to determine which data structures and applications are most improved by our use of multiple solvers (ignoring the time spend in the decision making process).
-This information could lead to future advancements both in application specific tasks (where the behavior of our solver selector could be manually replicated) and in generalized cases, as more efficient solver selectors / feature extractors are found.
+This information could lead to future advancements both in application specific tasks, where the behavior of our solver selector could be manually replicated, and in generalized cases, as more efficient solver selectors / feature extractors are found.
 
@@
 
-If one were to think of the most naive possible way to do this, you could imagine randomly sampling the bag of solvers (including their hyperparameters) and the training data to build a table of (feature set, solver) $\rightarrow$ cost.
-One could then perform a variant of nearest neighbor on that table given an input $x$ to find the solver with the lowest cost.
+The most naive possible way to do this would involve randomly sampling the bag of solvers (including their hyperparameters) and the training data to build a table of (feature set, solver) $\rightarrow$ cost.
+We could then perform a variant of nearest neighbor on that table given an input $x$ to find the solver with the lowest cost.
 
 However, this approach is rife with complication.
-First, it is difficult using this framework to identify relationships between solvers (for example, if one solver is very good at solving its own subproblems).
-It would even be difficult to individually measure solver performance, as a fast solver that calls big subproblems might actually be less efficient than a slow solver that calls small subproblems.
-Even if one tries to get around this issue by averaging subproblem approaches, one still runs into the above issue of failing to model more complex relationships between solvers and the problems they best perform on.
+First, it is difficult to individually measure solver performance, as a fast solver that calls big subproblems might actually be less efficient than a slow solver that calls small subproblems.
+Even if we try to get around this issue by averaging subproblem approaches, we would fail to model more complex relationships between solvers (for example, if one solver is particularly good at solving its own subproblems).
 
@@
 As LHYRA is partially modelled after RMI, we expect this to be a fairly straightforward adaptation.
-	\item \textbf{Sorting.} We’d like to show how LHYRA can adapt to different distributions in choosing the optimal sorting method.
+Additionally, LHYRA should make it easier to experiment with additional CDF predictors.
+	\item \textbf{Sorting.} We’'d like to show how LHYRA can adapt to different distributions in choosing the optimal sorting method.
 We’d also like to show that LHYRA can also adapt to quickly sort approximately, and utilize processor parallelism where available.
@@
 This is a place where LHYRA may be invaluable, and also has clear real-world applications.
-	\item \textbf{Matrix Multiplication.} LHYRA should be able to adaptively select naive matrix multiplication, naive divide and conquer, Strassen’s algorithm, or Coppersmith–Winograd.
+	\item \textbf{Matrix Multiplication.} LHYRA should be able to adaptively select naive matrix multiplication, naive divide and conquer, Strassen’s algorithm, or Coppersmith-–Winograd.
 This would be particularly useful for machine learning.
-	Nearest Neighbor: Given high-dimensional data, find the nearest neighbor. Examples of methods for this could be Locality sensitive hashing, Greedy search in proximity neighborhoods, and naive search.
+	\item \textbf{Nearest Neighbor.} Given high-dimensional data, find the nearest neighbor.
+Examples of methods for this could be Locality sensitive hashing, Greedy search in proximity neighborhoods, and naive (linear) search.
 
@@
 We intend to develop LHYRA in Python for the sake of implementation speed, as we can still demonstrate performance improvements over naive Python implementations.
+Additionally, if we have sufficiently promising results from the Python implementation, we may also port several of the generated solutions to C/C++ to see how LHYRA compares to the state of the art.
 

### 2019-05-16 c++/points.cpp
.

@@ -164,12 +164,11 @@ int main() {
 		&smart_opt
 	};
 
-	linopt.train(10,10,true);
+	linopt.train(80,20,true);
 
-
-	std::vector<Point> points; points.reserve(4000);
+;
 	std::ofstream f("data.txt");
-	const int NUM_DATA = 100;
+	const int NUM_DATA = 40;
 	for(LIST_LENGTH = 16; LIST_LENGTH <= 4096; LIST_LENGTH *= 2) {
 
 		std::cout << "Starting list of length " << LIST_LENGTH << std::endl;

### 2019-05-16 c++/sort_test.cpp

@@ -11,7 +11,8 @@
 
 
 // Data generator for sorting. Yes, you modify the globals to change the behavior. Sorry.
-int LIST_LENGTH = 1000;
+int LIST_LENGTH = 100;
+bool is_training = true;
 double NEARLY_SORTED = 0;
 void shuffle_slice(std::vector<double> & a, int start, int stop) {
 	int i = start;
@@ -25,18 +26,21 @@ void shuffle_slice(std::vector<double> & a, int start, int stop) {
 }
 std::vector<double> random_list() {
 	std::vector<double> list;
-	list.reserve(LIST_LENGTH);
-	for(int i = 0; i < LIST_LENGTH; i++) {
+	int length;
+	if(is_training) length = rand() % LIST_LENGTH;
+	else length = LIST_LENGTH;
+	list.reserve(length);
+	for(int i = 0; i < length; i++) {
 		list.emplace_back(i);
 	}
 	if(double(rand()) / RAND_MAX < NEARLY_SORTED) {
 		// Create 8 sections of length 5 which are shuffled:
 		for(int i = 0; i < 8; i++) {
-			int start = rand() % (LIST_LENGTH-5);
+			int start = rand() % (length-5);
 			shuffle_slice(list, start, start+5);
 		}
 	}
-	else shuffle_slice(list, 0, LIST_LENGTH-1);
+	else shuffle_slice(list, 0, length-1);
 	return list;
 }
 
@@ -129,7 +133,8 @@ int main() {
 
 	//std::cout << "Checkpoint 1" << std::endl;
 
-	linopt.train(200, 40, true);
+	linopt.train(200, 50, true);
+	is_training = false;
 
 	//std::cout << "Checkpoint " << std::endl;
 

### 2019-05-16 c++/tests.sh
figure updates

@@ -9,8 +9,8 @@ echo "Running sorting code..."
 
 echo "Generating plots..."
 
-python plot_progress.py
-mv progress.png progress_sorting.png
+python make_graph_training.py
+mv training_times.png progress_sorting.png
 python make_graph.py
 
 echo "Running points code..."
@@ -19,8 +19,8 @@ echo "Running points code..."
 
 echo "Generating plot..."
 
-python plot_progress.py
-mv progress.png progress_points.png
+python make_graph_training.py
+mv training_times.png progress_points.png
 python make_graph_points.py
 
 echo "Uploading to imgur..."

### 2019-05-16 c++/make_graph.py
Added two new graphs.

@@ -16,13 +16,11 @@ with open('data.txt', 'r') as f:
 		merge.append(log2(int(m)))
 		insertion.append(log2(int(i)))
 		quick.append(log2(int(q)))
-		std.append(log2(int(s)))
 
 plt.plot(xs, lhyra, label='lhyra')
 plt.plot(xs, merge, label='merge')
 plt.plot(xs, insertion, label='insertion')
 plt.plot(xs, quick, label='quick')
-plt.plot(xs, std, label='std')
 
 plt.xlabel('Log of List Length')
 plt.ylabel('Log of Time (in microseconds)')

### 2019-05-16 paper/main.tex
figure updates
all figures in paper!

@@
 
-\vspace{0.5cm}
 \begin{figure}[!h]
@@
         }{
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/training_times_points}
         }
@@
         }{%
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/relative_plot_points}
         }
@@
 
-\vspace{0.5cm}
 \begin{figure}[!ht]
@@
         }{
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/training_times_unsorted_lists}
         }
@@
         }{%
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/relative_plot_unsorted_lists}
         }
@@
 
-However, Lhyra can perform even better with interesting and changing data distributions.  Take the following example, where lists are already 50\% sorted.
+However, Lhyra can perform even better with interesting and changing data distributions.  Take the following example, where lists are 80\% nearly sorted.
 
-\vspace{0.5cm}
 \begin{figure}[!ht]
@@
         }{
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/training_times_mostly_sorted_lists}
         }
@@
         }{%
-            \includegraphics[width=5.5cm]{example-image}
+            \includegraphics[width=8cm]{images/relative_plot_mostly_sorted_lists}
         }

### 2019-05-16 paper/main.tex
added to discussion.

@@
 In this scenario, cost could be a weighted function of evaluation time and accuracy, depending on the priorities of the user.
+It is also likely that for these more difficult problems, the benefits of Lhyra could be far greater in magnitude than they are in \textsc{Sorting} and \textsc{Closest Pair}, because optimization for time on already-fast problems constrains the types of features and complexity of optimizer that Lhyra can use.
+For problems relating to protein-folding, for example, Lhyra could use more complex features, models, and solvers, and operate on more diverse data distributions, and possibly yiled big gains for a problem which is useful and hard.
+
+In conclusion, we have designed and implemented a general framework for learned optimization of recursive functions.
+We have shown that Lhyra can learn models for simple problems whih provide substantial improvements over individual algorithms, and also demonstrated the generality and extensibiity of our approach, with an eye towards future work.
+We believe Lhyra constitutes a useful development in the field of learning-augmented algorithms.
 

### 2019-05-16 paper/main.tex
paper formatting updates

@@
-\documentclass{article}
+\documentclass[11pt]{article}
 \usepackage{hyperref, graphicx, floatrow}
 
-\usepackage[letterpaper, margin=1in]{geometry}
+\usepackage[letterpaper, margin=1.25in]{geometry}
 \setlength{\parskip}{\baselineskip}
@@
 It is also likely that for these more difficult problems, the benefits of Lhyra could be far greater in magnitude than they are in \textsc{Sorting} and \textsc{Closest Pair}, because optimization for time on already-fast problems constrains the types of features and complexity of optimizer that Lhyra can use.
-For problems relating to protein-folding, for example, Lhyra could use more complex features, models, and solvers, and operate on more diverse data distributions, and possibly yiled big gains for a problem which is useful and hard.
+For problems relating to protein-folding, for example, Lhyra could use more complex features, models, and solvers, and operate on more diverse data distributions, and possibly yield big gains for a problem which is useful and hard.
 
